---
title: "Video Game Sales Analysis"
subtitle: "Industry trends across time, genres, and publishers"
author: "Florijan Peric"

format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true

execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This project analyzes the Video Game Sales dataset. The goal is to investigate:

- Which genres generate the most sales.
- How video game sales changed over time.
- Which publishers dominate the market.
- Regional differences in sales.

```{r}
library(tidyverse)
library(knitr)

games <- read_csv("vgsales.csv", show_col_types = FALSE)
```

# Data Presentation

## First Rows

```{r}
head(games)
```

## Structure

```{r}
glimpse(games)
```

## Column Description

- Rank: Sales ranking.
- Name: Game title.
- Platform: Gaming platform.
- Year: Release year.
- Genre: Game genre.
- Publisher: Publisher.
- NA_Sales: North America sales (millions).
- EU_Sales: Europe sales (millions).
- JP_Sales: Japan sales (millions).
- Other_Sales: Other regions.
- Global_Sales: Worldwide sales.

# Data Transformation

## User Defined Function

```{r}
top_n_sales <- function(data, n = 10) {
  data |>
    arrange(desc(Global_Sales)) |>
    slice_head(n = n)
}
```

## Top Selling Games

```{r}
top_games <- top_n_sales(games, 10)

kable(top_games)
```
Explanation: The table shows the 10 best-selling games.

## Sales by Genre

```{r}
genre_sales <- games |>
  group_by(Genre) |>
  summarise(
    Total_Sales = sum(Global_Sales, na.rm = TRUE)
  ) |>
  arrange(desc(Total_Sales))

kable(genre_sales)
```
Explanation: Data are aggregated by genre using group_by() and summarise().

## Top Publishers

```{r}
publisher_sales <- games |>
  group_by(Publisher) |>
  summarise(
    Total_Sales = sum(Global_Sales, na.rm = TRUE)
  ) |>
  arrange(desc(Total_Sales)) |>
  slice(1:15)

kable(publisher_sales)
```
Explanation: This table identifies the largest publishers.

## Sales by Year

```{r}
year_sales <- games |>
  filter(!is.na(Year)) |>
  group_by(Year) |>
  summarise(
    Total_Sales = sum(Global_Sales, na.rm = TRUE)
  )

kable(head(year_sales,20))
```
Explanation: Annual sales are calculated after filtering missing years.

## Join Example

```{r}
genre_count <- games |>
  group_by(Genre) |>
  summarise(Game_Count = n())

genre_join <- genre_sales |>
  inner_join(genre_count, by = "Genre")

kable(genre_join)
```
Explanation: inner_join() combines genre sales and genre counts.

# Exploratory Data Analysis

## Plot 1: Sales by Genre

```{r}
ggplot(
  genre_sales,
  aes(x = reorder(Genre, Total_Sales), y = Total_Sales)
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Global Sales by Genre",
    x = "Genre",
    y = "Sales (millions)"
  ) +
  theme_minimal()
```

Interpretation: Action and Sports genres are among the strongest categories.

## Plot 2: Sales Through Time

```{r}
year_sales <- year_sales %>%
  mutate(
    Year = trimws(Year),   # odstrani presledke
    Year = as.numeric(Year)
  ) %>%
  filter(!is.na(Year))

ggplot(year_sales,
       aes(x = Year, y = Total_Sales)) +
  geom_area(alpha = 0.4) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_x_continuous(
    breaks = seq(1980, 2020, by = 5)
  ) +
  labs(
    title = "Global Video Game Sales by Year",
    x = "Year",
    y = "Sales (millions)"
  ) +
  theme_minimal()
```

Interpretation: The graph shows the growth and later decline of the market.

## Plot 3: Top Publishers

```{r}
ggplot(
  publisher_sales,
  aes(x = reorder(Publisher, Total_Sales), y = Total_Sales)
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Publishers by Global Sales",
    x = "Publisher",
    y = "Sales"
  ) +
  theme_minimal()
```

Interpretation: A few major publishers dominate the market.

## Plot 4: Regional Sales

```{r}
region_sales <- games |>
  summarise(
    North_America = sum(NA_Sales, na.rm = TRUE),
    Europe = sum(EU_Sales, na.rm = TRUE),
    Japan = sum(JP_Sales, na.rm = TRUE),
    Other = sum(Other_Sales, na.rm = TRUE)
  ) |>
  pivot_longer(
    everything(),
    names_to = "Region",
    values_to = "Sales"
  )

ggplot(region_sales,
       aes(x = Region, y = Sales)) +
  geom_col() +
  labs(
    title = "Regional Sales Comparison",
    x = "Region",
    y = "Sales"
  ) +
  theme_minimal()
```

Interpretation: North America contributes the largest share of sales.

# Conclusion

The analysis shows that:

1. Action and Sports are among the most successful genres.
2. A small number of publishers account for a large portion of total sales.
3. Global sales increased strongly before reaching a peak and later declining.
4. North America is the largest market for video games.

The project demonstrates the use of tidyverse tools including filter(), mutate(), summarise(), inner_join(), ggplot(), and user-defined functions.

# References

- Kaggle Video Game Sales Dataset
- https://www.kaggle.com/datasets/gregorut/videogamesales
- Tidyverse Documentation
- ggplot2 Documentation
